---
title: "Power and Sample Size Calculation Tools"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{examples}
  %\VignetteEngine{knitr::rmarkdown}
  \usepackage[utf8]{inputenc}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(pwrss)
```

To install and load `pwrss`:
```{r,  message = FALSE, eval = FALSE}
install.packages("pwrss")
library(pwrss)
```

`pwrss` R package allows statistical power and minimum required sample size calculations for 

  - `(1)` testing a proportion (one-sample) against a constant, 
  - `(2)` testing a mean (one-sample) against a constant,  
  - `(3)` testing difference between two proportions (independent samples),
  - `(4)` testing difference between two means (independent and paired samples),
  - `(5)` testing a correlation (one-sample) against a constant,  
  - `(6)` testing difference between two correlations (independent samples),
  - `(7)` testing an R-squared against zero in linear regression 
  - `(8)` testing an R-squared difference against zero in hierarchical regression,
  - `(9)` testing an eta-squared or f-squared (for main and interaction effects) against zero in analysis of variance (ANOVA) (could be one-way, two-way, and three-way), 
  - `(10)` testing an eta-squared or f-squared (for main and interaction effects) against zero in analysis of covariance (ANCOVA) (could be one-way, two-way, and three-way), 
  - `(11)` testing an eta-squared or f-squared (for between, within, and interaction effects) against zero in one-way repeated measures analysis of variance (RM-ANOVA) (with non-sphericity correction and repeated measures correlation). 
  
  Alternative hypothesis can be formulated as "not equal", "less", "greater",  "non-inferior", "superior", or "equivalent" in `(1)`, `(2)`, `(3)`, and `(4)`;
  as "not equal", "less", or "greater" in `(5)` and `(6)`; but always as "greater" in `(7)`, `(8)`, `(9)`, `(10)`, and `(11)`. 
  
**If you find the package and related material useful, along with other references, please cite as:**

- Bulus, M. (2022). pwrss: Power and Sample Size Calculation Tools. R package version 0.1.0. https://CRAN.R-project.org/package=pwrss

- Bulus, M., & Polat, C. (2022, November 7). Istatistiksel Guc Analizine Giris [Introduction to Statistical Power Analysis]. https://doi.org/10.35542/osf.io/tfyxq 

**A Few Snapshots** 
  
**1. Generic Functions for _z_, _t_, and _F_ tests**

These generic functions calculate and return statistical power along with optional plots for Type I and Type II error rates (as long as the test statistic and degrees of freedom is known). They can be useful for calculating statistical power after data analysis (ex-post). 
```{r, message = FALSE, fig.width = 7, fig.height = 5, results = FALSE}
power.t.test(ncp = 1.96, df = 99, alpha = 0.05, alternative = "equivalent")
power.z.test(ncp = 1.96, alpha = 0.05, alternative = "equivalent")
power.f.test(ncp = 2, df1 = 2, df2 = 98, alpha = 0.05)

```

`plot()` function (S3 method) is a wrapper around the generic functions above. 
```{r, message = FALSE, fig.width = 7, fig.height = 5, results = FALSE}
design1 <- pwrss.t.2means(mu1 = 0.20, margin = -0.05, paired = TRUE,
                         power = 0.80, alternative = "non-inferior")
plot(design1)

design2 <- pwrss.f.ancova(eta2 = 0.10, n.levels = c(2,3),
                         power = .80)
plot(design2)

```

**2. Standardizd versus Unstandardized Input for _t_ Test**
```{r, message = FALSE, fig.width = 7, fig.height = 5, results = TRUE}
# it is sufficient to provide standardized difference between two groups for 'mu1'
# e.g. Cohen's d = 0.50 for mu1, because mu2 = 0 by default
pwrss.t.2means(mu1 = 0.50, power = .80)
pwrss.t.2means(mu1 = 0.50, power = .80, paired  = TRUE)

# it is sufficient to provide pooled standard deviation for sd1
# because sd2 = sd1 by default
pwrss.t.2means(mu1 = 10, mu2 = 5, sd1 = 10, power = .80)
```

**3. One-way, Two-way, and Three-way Analysis of Variance (ANOVA) or Analysis of Covariance (ANCOVA)**

```{r, message = FALSE, fig.width = 7, fig.height = 5, results = TRUE}

# one-way ANOVA
pwrss.f.ancova(n.levels = 2, n.covariates = 0,
               power = .80, eta2 = 0.10)
# one-way ANCOVA
pwrss.f.ancova(n.levels = 2, n.covariates = 1,
               power = .80, eta2 = 0.08)

# two-way ANOVA
pwrss.f.ancova(n.levels = c(2,3), n.covariates = 0,
               power = .80, eta2 = 0.10)
# two-way ANCOVA
pwrss.f.ancova(n.levels = c(2,3), n.covariates = 1,
               power = .80, eta2 = 0.08)

# three-way ANOVA
pwrss.f.ancova(n.levels = c(2,3,2), n.covariates = 0,
               power = .80, eta2 = 0.10)
# three-way ANCOVA
pwrss.f.ancova(n.levels = c(2,3,2), n.covariates = 1,
               power = .80, eta2 = 0.08)

```

**4. One-way Repeated Measures Analysis of Variance (RM-ANOVA)**

```{r, message = FALSE, fig.width = 7, fig.height = 5, results = TRUE}

pwrss.f.rmanova(eta2 = 0.10,  n.levels = 2, n.measurements = 3,
                 repmeasures.r = 0.50, type = "between", power = 0.80)

pwrss.f.rmanova(eta2 = 0.10,  n.levels = 2, n.measurements = 3,
                repmeasures.r = 0.50, epsilon = 1,
                type = "within", power = 0.80)

pwrss.f.rmanova(eta2 = 0.10,  n.levels = 2, n.measurements = 3,
                repmeasures.r = 0.50, epsilon = 1,
                type = "interaction", power = 0.80)

```

--o--
